The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-12-01 article_final/main.tex
Edited with Grammarly

Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam \citep{kingma2014adam} and AdamW \citep{loshchilov2017decoupled}, recently proposed Muon \citep{jordan2024muon} has shown promising results on training very large models \citep{liu2025muon}. Its key difference from Adam and AdamW is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. After 1750 iterations on NanoGPT training, F-Muon achieves 3.281 cross-entropy loss, while Muon achieves 3.279. Similarly, after 5960 iterations of GPT-2 Medium training we get 2.9215 with F-Muon and 2.9198 with Muon. Finally, S-Muon matches Muon on fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\mU\mV^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX^{t+1}) := F(\mX^t) + \<\mM^t, \mX^{t+1} - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

Unlike Schatten norms, which satisfy $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$ (for $p^{-1} + q^{-1} = 1$) and are thus closed under dualization, Ky Fan norms are generally not. Only two exceptional cases exist: the dual to the Ky Fan $1$-norm $\norm{\cdot}_{\text{KF-}1}$ (the spectral norm) is the Ky Fan $\min(m,n)$-norm $\norm{\cdot}_{\text{KF-}\min(m,n)}$ (the nuclear norm), and vice versa. Thus working with duals of Ky Fan norms can open up new possibilities:

@@

%Simply applying norms dual to Ky Fan $k$-norms, however, does not provide sufficient algorithmic diversity for our purposes.
The approaches to design new LMO algorithms are not limited to applying norms dual to Ky Fan $k$-norms. We now consider linear combinations of LMO-based algorithms and show that these combinations are themselves LMO-algorithms.

@@

We evaluate several algorithms from the Fanion family and their convex combinations, using the update rule from \equationref{eq:simple_update} with approximate Nesterov momentum:
\begin{itemize}
  \item \textbf{Fanions:} Neon (Fanion-1), Fanion-2, Fanion-10, Fanion-100, and Muon (Fanion-500).
  \item \textbf{F-Fanions:} F-Neon (F-Fanion-1), F-Fanion-2, F-Fanion-10, F-Fanion-100, and F-Muon (F-Fanion-500) with $\alpha=1/2$.
  \item \textbf{S-Fanions:} S-Neon (S-Fanion-1), S-Fanion-2, S-Fanion-10, S-Fanion-100, and S-Muon (S-Fanion-500) with $\alpha=1/2$ and \texttt{sign\_lr\_coeff=0.01}.
\end{itemize}
We also include baselines Normalized SGD and SignSGD, which correspond to F-Fanion and S-Fanion respectively with $\alpha = 0$ and arbitrary $k$.

@@

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [5 \cdot 10^{-5},\, 2 \cdot 10^{-4}]$ with step $1 \cdot 10^{-5}$.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

@@

\begin{figure}[htbp]
\floatconts
  {fig:app_lls}
  {\caption{More plots for linear least squares problem for a 500x500 matrix.}}
  {%
    \subfigure[The spectral norm of the full gradient]{\label{fig:lls_op_grad_norm}%
      \includegraphics[width=0.49\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
    \hfill
    \subfigure[The nuclear norm of the full gradient]{\label{fig:lls_nuclear_grad_norm}%
      \includegraphics[width=0.49\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
    \vspace{1em}
    \subfigure[The loss over time]{\label{fig:lls_loss_time}%
      \includegraphics[width=0.49\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
  }
\end{figure}

@@

A uniform training protocol was applied to all runs, including identical batch size, warm-up and cosine decay learning rate schedule, gradient clipping strategy, and validation split. Our optimization methods were integrated into the NanoGPT training loop without modifying the model architecture or preprocessing pipeline. Model quality was primarily evaluated using validation loss tracked over 200 training steps.

>>>>

 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part of training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam \citep{kingma2014adam} and AdamW \citep{loshchilov2017decoupled}, recently proposed Muon \citep{jordan2024muon} has shown promising results on training very large models \citep{liu2025muon}. Its key difference from Adam and AdamW is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

 Most notably, our experiments on real-world tasks demonstrate that the choice of underlying matrix norm is remarkably flexible. On CIFAR-10 airbench, properly-tuned F-Muon and S-Muon achieve $94.02 \pm 0.13\%$ and $94.03 \pm 0.13\%$ accuracy, matching Muon's $94.01 \pm 0.13\%$ performance. After 1750 iterations on NanoGPT training, F-Muon achieves 3.281 cross-entropy loss, while Muon achieves 3.279. Similarly, after 5960 iterations of GPT-2 Medium training, we get 2.9215 with F-Muon and 2.9198 with Muon. Finally, S-Muon matches Muon in fine-tuning NanoGPT on Tiny Stories, while F-Muon is far more resistant to learning rate choice than Muon. These results show that Muon-like algorithms can maintain competitive performance even when the underlying norm constraint is significantly modified, providing an affirmative answer to the central question posed above. Moreover, the tools from \sectionref{sec:conic_combo} give researchers considerable flexibility in designing algorithms that need not be direct modifications of Muon.

@@

 To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
 \[
     \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
 \]
 According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
 \[
 \<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
 \]
 The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\mU\mV^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
 \[
 \mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
 \]
 depending on which of those two expressions achieves a smaller residual on the linear function $L(\mX^{t+1}) := F(\mX^t) + \<\mM^t, \mX^{t+1} - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

 Unlike Schatten norms, which satisfy $\norm{\mM^t}_{S_p}^\dagger = \norm{\mM^t}_{S_q}$ (for $p^{-1} + q^{-1} = 1$) and are thus closed under dualization, Ky Fan norms are generally not. Only two exceptional cases exist: the dual to the Ky Fan $1$-norm $\norm{\cdot}_{\text{KF-}1}$ (the spectral norm) is the Ky Fan $\min(m,n)$-norm $\norm{\cdot}_{\text{KF-}\min(m,n)}$ (the nuclear norm), and vice versa. Thus, working with duals of Ky Fan norms can open up new possibilities:

@@

 %Simply applying norms dual to Ky Fan $k$-norms, however, does not provide sufficient algorithmic diversity for our purposes.
 The approaches to designing new LMO algorithms are not limited to applying norms dual to Ky Fan $k$-norms. We now consider linear combinations of LMO-based algorithms and show that these combinations are themselves LMO-algorithms.

@@

 We evaluate several algorithms from the Fanion family and their convex combinations, using the update rule from \equationref{eq:simple_update} with approximate Nesterov momentum:
 \begin{itemize}
   \item \textbf{Fanions:} Neon (Fanion-1), Fanion-2, Fanion-10, Fanion-100, and Muon (Fanion-500).
   \item \textbf{F-Fanions:} F-Neon (F-Fanion-1), F-Fanion-2, F-Fanion-10, F-Fanion-100, and F-Muon (F-Fanion-500) with $\alpha=1/2$.
   \item \textbf{S-Fanions:} S-Neon (S-Fanion-1), S-Fanion-2, S-Fanion-10, S-Fanion-100, and S-Muon (S-Fanion-500) with $\alpha=1/2$ and \texttt{sign\_lr\_coeff=0.01}.
 \end{itemize}
 We also include baselines Normalized SGD and SignSGD, which correspond to F-Fanion and S-Fanion, respectively, with $\alpha = 0$ and arbitrary $k$.

@@

 We perform a grid search over the following hyperparameter ranges:
 \begin{itemize}
   \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
   \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
   \item \texttt{lr} $\in [5 \cdot 10^{-5},\, 2 \cdot 10^{-4}]$ with step $1 \cdot 10^{-5}$.
   \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
 \end{itemize}
 The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon, respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

@@

 \begin{figure}[htbp]
 \floatconts
   {fig:app_lls}
   {\caption{More plots for the linear least squares problem for a 500x500 matrix.}}
   {%
     \subfigure[The spectral norm of the full gradient]{\label{fig:lls_op_grad_norm}%
       \includegraphics[width=0.49\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_long_run.pdf}}%
     \hfill
     \subfigure[The nuclear norm of the full gradient]{\label{fig:lls_nuclear_grad_norm}%
       \includegraphics[width=0.49\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_long_run.pdf}}%
     \vspace{1em}
     \subfigure[The loss over time]{\label{fig:lls_loss_time}%
       \includegraphics[width=0.49\linewidth]{figs/simple_lls/loss_vs_time_long_run.pdf}}
   }
 \end{figure}

@@

 A uniform training protocol was applied to all runs, including an identical batch size, warm-up and cosine decay learning rate schedule, gradient clipping strategy, and validation split. Our optimization methods were integrated into the NanoGPT training loop without modifying the model architecture or preprocessing pipeline. Model quality was primarily evaluated using validation loss tracked over 200 training steps.
